---
title: "Colorado Retail Ammunition Market Analysis"
author: "Colorado Lead Free Analysis Team"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
subtitle: "Exploratory Pricing, Availability, and Lead vs. Lead-Free Comparisons"
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 6,
  dpi = 300
)

library(tidyverse)
library(knitr)
library(scales)
library(readxl)
library(writexl)

# Focus on common big game calibers in Colorado/Western hunting
key_calibers <- c("270 Win", "308 Win", "30-06 Springfield", "6.5 Creedmoor", "300 Win Mag", "300 WSM", "243 Win", "7mm Rem Mag", "7mm-08 Rem")

# Load cleaned dataset
ammo <- read_excel("results/ammo_clean.xlsx") %>%
  filter(
    !str_detect(coalesce(name, ""), regex("FMJ|Full Metal", ignore_case = TRUE))
  ) %>%
  mutate(dedup_id = coalesce(upc, mpn, name)) %>%
  group_by(retailer, dedup_id) %>%
  slice_max(order_by = date, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  filter(caliber %in% key_calibers)
```

\newpage

## Methods & Data Processing

### Data Collection & Scope
* **Data Sources:** Pricing and catalog data were aggregated across major brick-and-mortar storefronts in Colorado and online retailers shipping directly into the state.
* **Regulation Alignment:** Full Metal Jacket (FMJ), match target ammunition, and solid non-expanding target rounds were filtered out to reflect projectiles legally permitted for big-game harvest under Colorado Parks and Wildlife (CPW) regulations (expanding/mushrooming bullets only).
* **Caliber Focus:** Analysis is restricted to a sampling of popular Colorado big-game centerfire rifle calibers, excluding rimfire, pistol cartridges, and bulk military-surplus chamberings.

### Deduplication & Standardization
* **Product-Level Identity:** Individual listings were deduplicated across retailers using Universal Product Codes (UPC) and Manufacturer Part Numbers (MPN), falling back to title matching where identifiers were unlisted.
* **Temporal Freshness:** For retailers with repeat scrapes over time, only the most recent observed price per product was retained.
* **Pricing Unit:** All costs reflect factory loaded ammunition normalized to Price Per Round (USD/rd). Box counts (20-round standard hunting boxes vs. 50-round packages) were standardized by dividing total box price by box quantity prior to aggregation.

## Dataset Overview

```{r overall-summary}
overall_summary <- ammo %>%
  filter(!is.na(ppr), ppr > 0) %>%
  summarise(
    `Total Listings` = n(),
    `Mean PPR`       = mean(ppr, na.rm = TRUE),
    `Median PPR`     = median(ppr, na.rm = TRUE),
    `Min PPR`        = min(ppr, na.rm = TRUE),
    `Max PPR`        = max(ppr, na.rm = TRUE),
    `IQR PPR`        = IQR(ppr, na.rm = TRUE)
  )

kable(overall_summary, digits = 2, caption = "High-Level Market Summary (USD per Round)")
```

Across all tracked calibers and retailers, the typical price per round sits at `r scales::dollar(overall_summary[["Median PPR"]])` - the median listed above. Because high-end rounds can pull the arithmetic mean upward, this report primarily relies on the **median** (the middle price on the shelf) to represent where everyday consumer costs likely sit.

## Price Comparison: Lead vs. Lead-Free

```{r price-summary}
lead_vs_free_summary <- ammo %>%
  filter(!is.na(category), !is.na(caliber), !is.na(ppr)) %>%
  group_by(caliber, category) %>%
  summarise(
    count      = n(),
    median_ppr = median(ppr, na.rm = TRUE),
    mean_ppr   = mean(ppr, na.rm = TRUE),
    sd_ppr     = sd(ppr, na.rm = TRUE),
    .groups    = "drop"
  ) %>%
  arrange(caliber, category)

plot_data <- lead_vs_free_summary %>%
  select(caliber, category, median_ppr) %>%
  pivot_wider(names_from = category, values_from = median_ppr) %>%
  rename(Lead_Core = `Lead-Core`, Lead_Free = `Lead-Free`) %>%
  mutate(
    premium = Lead_Free - Lead_Core,
    caliber = reorder(caliber, Lead_Free)
  )
```

### Median Pricing Dumbbell Chart

```{r dumbbell-plot}
p_dumbbell <- ggplot(plot_data) +
  geom_segment(
    aes(x = Lead_Core, xend = Lead_Free, y = caliber, yend = caliber),
    color = "#a0a0a0", size = 1.2
  ) +
  geom_point(aes(x = Lead_Core, y = caliber, color = "Lead-Core"), size = 3) +
  geom_point(aes(x = Lead_Free, y = caliber, color = "Lead-Free"), size = 3) +
  geom_text(
    aes(
      x = pmax(Lead_Free, Lead_Core, na.rm = TRUE) + 0.1,
      y = caliber,
      label = ifelse(premium >= 0, sprintf("+$%.2f", premium), sprintf("-$%.2f", abs(premium)))
    ),
    hjust = 0, size = 3.5, color = "#2c3e50"
  ) +
  scale_color_manual(
    name = "Ammunition Type",
    values = c("Lead-Core" = "#2b5c8f", "Lead-Free" = "#d95f02")
  ) +
  scale_x_continuous(
    labels = dollar_format(),
    expand = expansion(mult = c(0.02, 0.12))
  ) +
  labs(
    title = "Median Price per Round: Lead-Core vs. Lead-Free",
    subtitle = "Comparing factory ammunition prices across common big game calibers",
    x = "Median Price per Round (USD)",
    y = NULL,
    caption = "Source: Scraped Retail Ammunition Data From Major Colorado Retailers"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "#eeeeee", linetype = "dashed"),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.text.y = element_text(face = "bold", size = 9),
    plot.caption = element_text(size = 8)
  )

p_dumbbell
```

**How to read this chart:**

- Each horizontal line represents a single caliber.
- The **blue dot** marks the median price of traditional lead-core rounds.
- The **orange dot** marks the median price of lead-free (copper/monolithic) rounds.
- The grey line between the two dots illustrates the price gap, with the dollar figure at the end indicating the median price difference per round.

**Takeaway:** Lead-free ammunition carries a consistent price surcharge across standard hunting calibers - generally between **+\$0.35 and +\$0.50 per round** (or roughly \$7.00 to \$10.00 more per 20-round box). In magnum cartridges (7mm Rem Mag, 300 Win Mag, .300 WSM), premium lead-core hunting loads approach price parity with lead-free offerings, sitting at a higher initial price per round in general.

\newpage

### Price Table

```{r price-table}
formatted_table <- plot_data %>%
  arrange(desc(Lead_Free)) %>%
  mutate(
    Lead_Core = sprintf("$%.2f", Lead_Core),
    Lead_Free = sprintf("$%.2f", Lead_Free),
    premium   = ifelse(premium >= 0, sprintf("+$%.2f", premium), sprintf("-$%.2f", abs(premium)))
  ) %>%
  rename(
    Caliber = caliber,
    `Lead-Core (\\$/rd)` = Lead_Core,
    `Lead-Free (\\$/rd)` = Lead_Free,
    `Lead-Free Premium` = premium
  )

kable(
  formatted_table,
  align = c("l", "r", "r", "r"),
  longtable = FALSE,
  caption = "Ammunition Pricing Comparison by Caliber")
```

The table highlights that entry costs depend heavily on the caliber itself. Historically popular cartridges like .243 Win, .308 Win, .270 and .30-06 Springfield - as well as more recent offerings including 6.5 Creedmoor - remain the most affordable overall, with the median price of lead-free loads around $2.50 per round.

## Market Availability & Distinct Offerings

```{r deduplication}
# Deduplicate on distinct products across all stores
unique_ammo <- ammo %>%
  filter(
    !str_detect(coalesce(name, ""), regex("FMJ|Full Metal", ignore_case = TRUE))
  ) %>%
  mutate(dedup_id = coalesce(upc, mpn, name)) %>%
  distinct(dedup_id, .keep_all = TRUE)

availability_data <- unique_ammo %>%
  group_by(caliber, category) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(caliber) %>%
  mutate(total_count = sum(count)) %>%
  ungroup() %>%
  mutate(caliber = reorder(caliber, total_count))
```

### Overall Distinct Products Available

```{r availability-plot}
p_availability <- ggplot(availability_data, aes(x = count, y = caliber, fill = category)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(
    aes(label = ifelse(count > 10, count, "")),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 3.5
  ) +
  scale_fill_manual(
    name = "Ammunition Type",
    values = c("Lead-Core" = "#2b5c8f", "Lead-Free" = "#d95f02")
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Market Availability: Distinct Products by Caliber",
    subtitle = "Unique hunting round product offerings (deduplicated across retailers)",
    x = "Distinct Products Available",
    y = NULL,
    caption = "Source: Scraped Retail Ammunition Data From Major Colorado Retailers"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(face = "bold")
  )

p_availability
```

**How to read this chart:**

* **Bar Length:** The total length shows the approximate number of distinct factory choices a consumer has for that caliber.
  * *Note on data:* While offerings have been deduplicated using universal product codes (UPC) and manufacturer part numbers (MPN), minor overlap may persist due to variations in how individual retailers catalog products.
* **Color Split:** Each bar is divided into **Lead-Core (blue)** and **Lead-Free (orange)** segments, illustrating the relative product diversity between the two bullet types.

**Takeaway:** Choice is heavily concentrated in a handful of calibers. The .308 Win, 6.5 Creedmoor, and .30-06 Springfield calibers represent the vast majority of retail variety. While lead-free options are available in every key hunting caliber, they generally represent 15% to 25% of total product variety, with some calibers like .300 WSM and 7mm-08 Rem offering fewer than 15 unique non-lead loads.

## Retailer-Level Availability

This section evaluates availability across individual storefronts - showing whether a hunter walking into a specific store or ordering from a specific retailer can reasonably expect to find lead-free choices.

```{r retailer-availability-table}
retailer_caliber_counts <- ammo %>%
  filter(!is.na(retailer), !is.na(caliber)) %>%
  group_by(retailer, caliber, category) %>%
  summarise(options_count = n(), .groups = "drop")

retailer_benchmark <- retailer_caliber_counts %>%
  group_by(caliber, category) %>%
  summarise(
    avg_options_per_retailer    = mean(options_count),
    median_options_per_retailer = median(options_count),
    retailers_stocking          = n(),
    .groups = "drop"
  ) %>%
  arrange(caliber, category)

kable(
  retailer_benchmark,
  digits = 1,
  caption = "Average & Median Product Options Available Per Retailer"
)
```

### Availability: Lead vs. Lead-Free by Retailer

```{r retailer-breadth-plot, fig.height=7}
retailer_caliber_counts %>%
  ggplot(aes(x = retailer, y = options_count, fill = category)) +
  geom_col(position = position_dodge(preserve = "single"), width = 0.7) +
  facet_wrap(~ caliber, scales = "free_y") +
  scale_fill_manual(
    name = "Ammunition Type",
    values = c("Lead-Core" = "#2b5c8f", "Lead-Free" = "#d95f02")
  ) +
  labs(
    title = "Distinct Product Selection by Retailer",
    subtitle = "Number of hunting round options carried per retailer for primary big game calibers",
    x = NULL,
    y = "Number of Distinct Options",
    caption = "Source: Scraped Retail Ammunition Data From Major Colorado Retailers"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )
```

**How to read this chart:**

- Each panel represents one caliber.
- The x-axis lists the individual retailers surveyed.
- The height of the bars shows how many distinct boxes of ammunition that retailer has on offer for that caliber, directly comparing **traditional lead (blue)** against **lead-free (orange)**.

**Takeaway:** Retail depth varies widely. While larger sporting-goods chains and online-retailers stock broad selections of lead-core options (often 10 to 30 choices per caliber), lead-free availability at the store level is frequently limited to 1 to 4 SKUs per caliber, meaning hunters shopping locally have substantially fewer projectile weights and bullet designs to choose from.

## Price Distributions: Where Do Shelf Prices Cluster?

Summary averages don't show whether options exist at budget price tiers or whether all lead-free offerings sit in the premium price bracket. The charts below display the full pricing spread across retailers.

### Price Spread & Range (Box-and-Whisker)

```{r box-whisker-plot, fig.height=8}
boxplot_data <- ammo %>%
  filter(
    !is.na(category),
    !is.na(ppr),
    ppr > 0,
    ppr <= 7.00,
    caliber %in% key_calibers
  )

p_boxplot <- ggplot(boxplot_data, aes(x = caliber, y = ppr, fill = category)) +
  geom_point(
    aes(color = category),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    alpha = 0.25,
    size = 1.2,
    show.legend = FALSE
  ) +
  geom_boxplot(
    width = 0.4,
    position = position_dodge(0.75),
    alpha = 0.6,
    outlier.shape = NA
  ) +
  scale_fill_manual(
    name = "Ammunition Type",
    values = c("Lead-Core" = "#2b5c8f", "Lead-Free" = "#d95f02")
  ) +
  scale_color_manual(
    values = c("Lead-Core" = "#2b5c8f", "Lead-Free" = "#d95f02")
  ) +
  scale_y_continuous(labels = dollar_format()) +
  coord_flip() +
  labs(
    title = "Price per Round Distribution by Caliber",
    subtitle = "Comparing median prices, middle-50% price ranges, and inventory spread",
    x = NULL,
    y = "Price per Round (USD)",
    caption = "Source: Scraped Retail Ammunition Data From Major Colorado Retailers"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(face = "bold")
  )

p_boxplot
```

**How to read this chart:**

- **The shaded box:** Covers the middle 50% of ammo prices (the interquartile range). If a box is short, prices are very uniform; if it is long, prices vary widely.
- **The solid line inside the box:** The median price point.
- **The "whiskers" (lines extending outward):** Show the typical upper and lower price limits for that ammunition.
- **The faint dots:** Represent individual boxes of ammunition, showing where retailer inventory is thin or dense.

**Takeaway:** The entry price for lead-free ammunition is noticeably higher than lead-core. While traditional lead ammunition features inexpensive baseline hunting options starting around \$1.25 to \$1.75 per round, lead-free rarely dips below \$2.10 to \$2.25 per round. At the top end, premium lead-core loads (tipped, bonded, high BC) overlap heavily with standard lead-free offerings.

### Price Density Profiles

```{r price-density-plot, fig.height=8}
density_data <- ammo %>%
  filter(
    !is.na(category),
    !is.na(ppr),
    ppr > 0,
    ppr <= 7.00,
    caliber %in% key_calibers
  )

ggplot(density_data, aes(x = ppr, fill = category, color = category)) +
  geom_density(alpha = 0.4, adjust = 1.2) +
  facet_wrap(~ caliber, scales = "free_y", ncol = 2) +
  scale_fill_manual(
    name = "Ammunition Type",
    values = c("Lead-Core" = "#2b5c8f", "Lead-Free" = "#d95f02")
  ) +
  scale_color_manual(
    name = "Ammunition Type",
    values = c("Lead-Core" = "#1e3f63", "Lead-Free" = "#a84900")
  ) +
  scale_x_continuous(labels = dollar_format()) +
  labs(
    title = "Price Density Profiles by Caliber",
    subtitle = "Overlay showing where factory ammunition prices cluster on the shelf",
    x = "Price per Round (USD)",
    y = "Concentration (Relative Density)",
    caption = "Source: Scraped Retail Ammunition Data From Major Colorado Retailers"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 14)
  )
```

**How to read this chart:** 

* Think of these curves as smooth hills showing **where products cluster along the price spectrum**.
* **High peaks:** Indicate a price point where many different products are available.
* **A curve shifted to the right:** Indicates that products in that category are systematically more expensive.
* **Overlapping shaded areas:** The price range where lead and lead-free options share the same price bracket.

**Takeaway:** Cartridges like .308 Win and 6.5 Creedmoor share a notable overlap under their curves, indicating that there are many lead-core and lead-free options sharing that price range. However, all calibers indicate that lead-free option are shifted toward an overall higher price per round. 
