---
title: "Colorado Retail Ammunition Market Analysis"
subtitle: "Exploratory Pricing, Availability, and Lead vs. Lead-Free Comparisons"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 6,
  dpi = 300
)

library(tidyverse)
library(knitr)
library(scales)
library(readxl)
library(writexl)

# Focus on common big game calibers in Colorado/Western hunting (e.g., 270 Win, 308 Win, 30-06, 6.5 Creedmoor)
key_calibers <- c("270 Win", "308 Win", "30-06 Springfield", "6.5 Creedmoor", "300 Win Mag", "300 WSM", "243 Win", "7mm Rem Mag", "7mm-08 Rem")

# Load cleaned dataset
ammo <- read_excel("results/ammo_clean.xlsx") %>%
  filter(
    !str_detect(coalesce(name, ""), regex("FMJ|Full Metal", ignore_case = TRUE))
  ) %>%
  mutate(dedup_id = coalesce(upc, mpn, name)) %>%
  group_by(retailer, dedup_id) %>%
  slice_max(order_by = date, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  filter(caliber %in% key_calibers)
  
  

#unique_ammo <- read_excel("results/unique_ammo_AG.xlsx", sheet = "Adam") %>%
#  filter(Exclude == FALSE)
```

## Dataset Overview

```{r overall-summary}
overall_summary <- ammo %>%
  filter(!is.na(ppr), ppr > 0) %>%
  summarise(
    `Total Listings` = n(),
    `Mean PPR`       = mean(ppr, na.rm = TRUE),
    `Median PPR`     = median(ppr, na.rm = TRUE),
    `Min PPR`        = min(ppr, na.rm = TRUE),
    `Max PPR`        = max(ppr, na.rm = TRUE),
    `IQR PPR`        = IQR(ppr, na.rm = TRUE)
  )

kable(overall_summary, digits = 2, caption = "High-Level Market Summary (USD per Round)")
```

---

## Price Comparison: Lead vs. Lead-Free

```{r price-summary}
lead_vs_free_summary <- ammo %>%
  filter(!is.na(category), !is.na(caliber), !is.na(ppr)) %>%
  group_by(caliber, category) %>%
  summarise(
    count      = n(),
    median_ppr = median(ppr, na.rm = TRUE),
    mean_ppr   = mean(ppr, na.rm = TRUE),
    sd_ppr     = sd(ppr, na.rm = TRUE),
    .groups    = "drop"
  ) %>%
  arrange(caliber, category)

plot_data <- lead_vs_free_summary %>%
  select(caliber, category, median_ppr) %>%
  pivot_wider(names_from = category, values_from = median_ppr) %>%
  rename(Lead_Core = `Lead-Core`, Lead_Free = `Lead-Free`) %>%
  mutate(
    premium = Lead_Free - Lead_Core,
    caliber = reorder(caliber, Lead_Free)
  )
```

### Median Pricing Dumbbell Chart

```{r dumbbell-plot}
p_dumbbell <- ggplot(plot_data) +
  geom_segment(
    aes(x = Lead_Core, xend = Lead_Free, y = caliber, yend = caliber),
    color = "#a0a0a0", size = 1.2
  ) +
  geom_point(aes(x = Lead_Core, y = caliber, color = "Lead-Core"), size = 3) +
  geom_point(aes(x = Lead_Free, y = caliber, color = "Lead-Free"), size = 3) +
  geom_text(
    aes(
      x = pmax(Lead_Free, Lead_Core, na.rm = TRUE) + 0.1,
      y = caliber,
      label = ifelse(premium >= 0, sprintf("+$%.2f", premium), sprintf("-$%.2f", abs(premium)))
    ),
    hjust = 0, size = 3.5, color = "#2c3e50"
  ) +
  scale_color_manual(
    name = "Ammunition Type",
    values = c("Lead-Core" = "#2b5c8f", "Lead-Free" = "#d95f02")
  ) +
  scale_x_continuous(
    labels = dollar_format(),
    expand = expansion(mult = c(0.02, 0.12))
  ) +
  labs(
    title = "Median Price per Round: Lead-Core vs. Lead-Free",
    subtitle = "Comparing factory ammunition prices across common big game calibers",
    x = "Median Price per Round (USD)",
    y = NULL,
    caption = "Source: Scraped Retail Ammunition Data From Major Colorado Retailers"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "#eeeeee", linetype = "dashed"),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.text.y = element_text(face = "bold", size = 9),
    plot.caption = element_text(size = 8)
  )

p_dumbbell
```

### Price Table

```{r price-table}
formatted_table <- plot_data %>%
  arrange(desc(Lead_Free)) %>%
  mutate(
    Lead_Core = sprintf("$%.2f", Lead_Core),
    Lead_Free = sprintf("$%.2f", Lead_Free),
    premium   = ifelse(premium >= 0, sprintf("+$%.2f", premium), sprintf("-$%.2f", abs(premium)))
  ) %>%
  rename(
    Caliber = caliber,
    `Lead-Core (\\$/rd)` = Lead_Core,
    `Lead-Free (\\$/rd)` = Lead_Free,
    `Lead-Free Premium` = premium
  )

kable(formatted_table, align = c("l", "r", "r", "r"), caption = "Ammunition Pricing Comparison by Caliber")
```

---

## Market Availability & Distinct Offerings

```{r deduplication}
# Deduplicate on distinct products excluding target/bulk FMJ
unique_ammo <- ammo %>%
  filter(
    !str_detect(coalesce(name, ""),  regex("FMJ|Full Metal", ignore_case = TRUE))
  ) %>%
  mutate(dedup_id = coalesce(upc, mpn, name)) %>%
  distinct(dedup_id, .keep_all = TRUE)

availability_data <- unique_ammo %>%
  group_by(caliber, category) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(caliber) %>%
  mutate(total_count = sum(count)) %>%
  ungroup() %>%
  mutate(caliber = reorder(caliber, total_count))
```

### Overall Distinct Products Available

```{r availability-plot}
p_availability <- ggplot(availability_data, aes(x = count, y = caliber, fill = category)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(
    aes(label = ifelse(count > 10, count, "")),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 3.5
  ) +
  scale_fill_manual(
    name = "Ammunition Type",
    values = c("Lead-Core" = "#2b5c8f", "Lead-Free" = "#d95f02")
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Market Availability: Distinct Products by Caliber",
    subtitle = "Unique hunting round product offerings (deduplicated by UPC/MPN/Name)",
    x = "Distinct Products Available",
    y = NULL,
    caption = "Source: Scraped Retail Ammunition Data From Major Colorado Retailers"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(face = "bold")
  )

p_availability
```

---

## Retailer-Level Availability

This explores the depth of inventory per storefront: how many options a consumer actually finds on shelves or online per retailer.

```{r retailer-availability-table}
# Calculate distinct offerings available per retailer by caliber and composition
retailer_caliber_counts <- ammo %>%
  filter(!is.na(retailer), !is.na(caliber)) %>%
  group_by(retailer, caliber, category) %>%
  summarise(options_count = n(), .groups = "drop")

# Average and median options carried across all retailers for each caliber
retailer_benchmark <- retailer_caliber_counts %>%
  group_by(caliber, category) %>%
  summarise(
    avg_options_per_retailer    = mean(options_count),
    median_options_per_retailer = median(options_count),
    retailers_stocking          = n(),
    .groups = "drop"
  ) %>%
  arrange(caliber, category)

kable(
  retailer_benchmark,
  digits = 1,
  caption = "Average & Median Product Options Available Per Retailer"
)
```

### Availability: Lead vs. Lead-Free by Retailer

```{r retailer-breadth-plot, fig.height=7}
# Focus on common big game calibers (e.g., 270 Win, 308 Win, 30-06, 6.5 Creedmoor)
key_calibers <- c("270 Win", "308 Win", "30-06 Springfield", "6.5 Creedmoor", "300 Win Mag", "300 WSM", "243 Win", "7mm Rem Mag", "7mm-08 Rem")

retailer_caliber_counts %>%
  ggplot(aes(x = retailer, y = options_count, fill = category)) +
  geom_col(position = position_dodge(preserve = "single"), width = 0.7) +
  facet_wrap(~ caliber, scales = "free_y") +
  scale_fill_manual(
    name = "Ammunition Type",
    values = c("Lead-Core" = "#2b5c8f", "Lead-Free" = "#d95f02")
  ) +
  labs(
    title = "Distinct Product Selection by Retailer",
    subtitle = "Number of hunting round options carried per retailer for primary big game calibers",
    x = NULL,
    y = "Number of Distinct Options",
    caption = "Source: Scraped Retail Ammunition Data From Major Colorado Retailers"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )
```

---

## Price Distribution & Density Curves

Overlaying density curves with transparency (`alpha = 0.45`) reveals price clustering, premium gaps, and distribution tails.

```{r price-density-plot, fig.height=8}
# Filter out extreme high-end outliers for cleaner visualization curves (e.g. ppr < $7.00)
density_data <- ammo %>%
  filter(
    !is.na(category),
    !is.na(ppr),
    ppr > 0,
    ppr <= 7.00,
    caliber %in% key_calibers
  )

ggplot(density_data, aes(x = ppr, fill = category, color = category)) +
  geom_density(alpha = 0.4, adjust = 1.2) +
  facet_wrap(~ caliber, scales = "free_y", ncol = 2) +
  scale_fill_manual(
    name = "Ammunition Type",
    values = c("Lead-Core" = "#2b5c8f", "Lead-Free" = "#d95f02")
  ) +
  scale_color_manual(
    name = "Ammunition Type",
    values = c("Lead-Core" = "#1e3f63", "Lead-Free" = "#a84900")
  ) +
  scale_x_continuous(labels = dollar_format()) +
  labs(
    title = "Price per Round Distribution by Caliber",
    subtitle = "Overlay density curves comparing Lead-Core vs. Lead-Free factory loads",
    x = "Price per Round (USD)",
    y = "Density",
    caption = "Source: Scraped Retail Ammunition Data From Major Colorado Retailers"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 14)
  )
```

```{r box-whisker-plot, fig.height=8}
library(scales)

# Filter for target calibers and reasonable bounds
boxplot_data <- ammo %>%
  filter(
    !is.na(category),
    !is.na(ppr),
    ppr > 0,
    ppr <= 7.00,
    caliber %in% key_calibers
  )

p_boxplot <- ggplot(boxplot_data, aes(x = caliber, y = ppr, fill = category)) +
  # Raw data points showing inventory volume/depth
  geom_point(
    aes(color = category),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    alpha = 0.25,
    size = 1.2,
    show.legend = FALSE
  ) +
  # Boxplot showing median, quartiles, and IQR
  geom_boxplot(
    width = 0.4,
    position = position_dodge(0.75),
    alpha = 0.6,
    outlier.shape = NA # Outliers are already visible via geom_point
  ) +
  scale_fill_manual(
    name = "Ammunition Type",
    values = c("Lead-Core" = "#2b5c8f", "Lead-Free" = "#d95f02")
  ) +
  scale_color_manual(
    values = c("Lead-Core" = "#2b5c8f", "Lead-Free" = "#d95f02")
  ) +
  scale_y_continuous(labels = dollar_format()) +
  coord_flip() + # Flip horizontal for legible caliber labels
  labs(
    title = "Price per Round Distribution by Caliber",
    subtitle = "Comparing medians, interquartile ranges (IQR), and product spreads",
    x = NULL,
    y = "Price per Round (USD)",
    caption = "Source: Scraped Retail Ammunition Data From Major Colorado Retailers"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(face = "bold")
  )

p_boxplot
```

